Reanalysis Step 2b: Multiple Imputation Outcome Diagnostics

### 2025-08-30 reanalysis-3-imputation_diagnostics.qmd
debug

```{r summary-stats}
# Compare change scores across all three groups
change_summary_detailed <- changes_df |>
  group_by(imputation_group) |>
  summarise(
    n = n(),
    mean_change = mean(change_baseline_to_outcome, na.rm = TRUE),
    sd_change = sd(change_baseline_to_outcome, na.rm = TRUE),
    median_change = median(change_baseline_to_outcome, na.rm = TRUE),
    q25_change = quantile(change_baseline_to_outcome, 0.25, na.rm = TRUE),
    q75_change = quantile(change_baseline_to_outcome, 0.75, na.rm = TRUE),
    min_change = min(change_baseline_to_outcome, na.rm = TRUE),
    max_change = max(change_baseline_to_outcome, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    iqr_change = q75_change - q25_change
  )

kable(change_summary_detailed |> 
        select(imputation_group, n, mean_change, sd_change, median_change, iqr_change, min_change, max_change),
      digits = 3,
      col.names = c("Group", "N", "Mean", "SD", "Median", "IQR", "Min", "Max"),
      caption = "Distribution of vision changes (outcome - baseline) by detailed imputation group")

# Also keep simplified version for compatibility
change_summary <- changes_df |>
  group_by(outcome_was_imputed) |>
  summarise(
    n = n(),
    mean_change = mean(change_baseline_to_outcome, na.rm = TRUE),
    sd_change = sd(change_baseline_to_outcome, na.rm = TRUE),
    median_change = median(change_baseline_to_outcome, na.rm = TRUE),
    q25_change = quantile(change_baseline_to_outcome, 0.25, na.rm = TRUE),
    q75_change = quantile(change_baseline_to_outcome, 0.75, na.rm = TRUE),
    min_change = min(change_baseline_to_outcome, na.rm = TRUE),
    max_change = max(change_baseline_to_outcome, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    iqr_change = q75_change - q25_change,
    group = ifelse(outcome_was_imputed, "Imputed", "Observed")
  )

# Statistical tests across three groups
anova_test <- aov(change_baseline_to_outcome ~ imputation_group, data = changes_df)
anova_summary <- summary(anova_test)

# Pairwise comparisons
pairwise_tests <- pairwise.t.test(changes_df$change_baseline_to_outcome, 
                                  changes_df$imputation_group,
                                  p.adjust.method = "bonferroni")

@@

# Format pairwise results
pairwise_matrix <- pairwise_tests$p.value
pairwise_df <- data.frame(
  Comparison = c("Imputed (with intermediate) vs Observed",
                "Imputed (no intermediate) vs Observed",
                "Imputed (no intermediate) vs Imputed (with intermediate)"),
  P_Value = c(pairwise_matrix[1,1], pairwise_matrix[2,1], pairwise_matrix[2,2]),
  Significant = ifelse(c(pairwise_matrix[1,1], pairwise_matrix[2,1], pairwise_matrix[2,2]) < 0.05,
                      "Yes", "No")
)

kable(pairwise_df, digits = 4, caption = "Pairwise comparisons (Bonferroni adjusted)")

# Also do simplified tests for backward compatibility
t_test <- t.test(change_baseline_to_outcome ~ outcome_was_imputed, data = changes_df)
ks_test <- ks.test(
  changes_df$change_baseline_to_outcome[changes_df$outcome_was_imputed],
  changes_df$change_baseline_to_outcome[!changes_df$outcome_was_imputed]
)
```

>>>>

```{r summary-stats}
# Compare change scores across all three groups
change_summary_detailed <- changes_df |>
  filter(!is.na(change_baseline_to_outcome)) |>  # Filter out NAs first
  group_by(imputation_group) |>
  summarise(
    n = n(),
    n_valid = sum(!is.na(change_baseline_to_outcome)),
    mean_change = if(all(is.na(change_baseline_to_outcome))) NA_real_ else mean(change_baseline_to_outcome, na.rm = TRUE),
    sd_change = if(all(is.na(change_baseline_to_outcome))) NA_real_ else sd(change_baseline_to_outcome, na.rm = TRUE),
    median_change = if(all(is.na(change_baseline_to_outcome))) NA_real_ else median(change_baseline_to_outcome, na.rm = TRUE),
    q25_change = if(all(is.na(change_baseline_to_outcome))) NA_real_ else quantile(change_baseline_to_outcome, 0.25, na.rm = TRUE),
    q75_change = if(all(is.na(change_baseline_to_outcome))) NA_real_ else quantile(change_baseline_to_outcome, 0.75, na.rm = TRUE),
    min_change = if(all(is.na(change_baseline_to_outcome))) NA_real_ else min(change_baseline_to_outcome, na.rm = TRUE),
    max_change = if(all(is.na(change_baseline_to_outcome))) NA_real_ else max(change_baseline_to_outcome, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    iqr_change = q75_change - q25_change
  )

# Only display if we have valid data
if(nrow(change_summary_detailed) > 0) {
  kable(change_summary_detailed |> 
          select(imputation_group, n, mean_change, sd_change, median_change, iqr_change, min_change, max_change),
        digits = 3,
        col.names = c("Group", "N", "Mean", "SD", "Median", "IQR", "Min", "Max"),
        caption = "Distribution of vision changes (outcome - baseline) by detailed imputation group")
}

# Also keep simplified version for compatibility
change_summary <- changes_df |>
  filter(!is.na(change_baseline_to_outcome)) |>  # Filter out NAs first
  group_by(outcome_was_imputed) |>
  summarise(
    n = n(),
    n_valid = sum(!is.na(change_baseline_to_outcome)),
    mean_change = if(all(is.na(change_baseline_to_outcome))) NA_real_ else mean(change_baseline_to_outcome, na.rm = TRUE),
    sd_change = if(all(is.na(change_baseline_to_outcome))) NA_real_ else sd(change_baseline_to_outcome, na.rm = TRUE),
    median_change = if(all(is.na(change_baseline_to_outcome))) NA_real_ else median(change_baseline_to_outcome, na.rm = TRUE),
    q25_change = if(all(is.na(change_baseline_to_outcome))) NA_real_ else quantile(change_baseline_to_outcome, 0.25, na.rm = TRUE),
    q75_change = if(all(is.na(change_baseline_to_outcome))) NA_real_ else quantile(change_baseline_to_outcome, 0.75, na.rm = TRUE),
    min_change = if(all(is.na(change_baseline_to_outcome))) NA_real_ else min(change_baseline_to_outcome, na.rm = TRUE),
    max_change = if(all(is.na(change_baseline_to_outcome))) NA_real_ else max(change_baseline_to_outcome, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    iqr_change = q75_change - q25_change,
    group = ifelse(outcome_was_imputed, "Imputed", "Observed")
  )

# Statistical tests across three groups - only if we have valid data
# Filter to non-missing values and check if we have enough groups
valid_data <- changes_df |>
  filter(!is.na(change_baseline_to_outcome) & !is.na(imputation_group))

# Check if we have at least 2 groups with data
groups_with_data <- valid_data |>
  group_by(imputation_group) |>
  summarise(n = n(), .groups = "drop") |>
  filter(n > 0)

if(nrow(groups_with_data) >= 2) {
  anova_test <- aov(change_baseline_to_outcome ~ imputation_group, data = valid_data)
  anova_summary <- summary(anova_test)
} else {
  anova_test <- NULL
  anova_summary <- NULL
}

# Pairwise comparisons - only if we have valid data
if(!is.null(anova_test) && nrow(groups_with_data) >= 2) {
  pairwise_tests <- pairwise.t.test(valid_data$change_baseline_to_outcome, 
                                    valid_data$imputation_group,
                                    p.adjust.method = "bonferroni")

@@

  # Format pairwise results if available
  if(!is.null(pairwise_tests$p.value)) {
    pairwise_matrix <- pairwise_tests$p.value

    # Build pairwise comparison dataframe dynamically based on available comparisons
    comparisons <- list()
    p_values <- list()

    # Check which comparisons are available
    if(nrow(pairwise_matrix) >= 1 && ncol(pairwise_matrix) >= 1 && !is.na(pairwise_matrix[1,1])) {
      comparisons <- c(comparisons, "Imputed (with intermediate) vs Observed")
      p_values <- c(p_values, pairwise_matrix[1,1])
    }
    if(nrow(pairwise_matrix) >= 2 && ncol(pairwise_matrix) >= 1 && !is.na(pairwise_matrix[2,1])) {
      comparisons <- c(comparisons, "Imputed (no intermediate) vs Observed")
      p_values <- c(p_values, pairwise_matrix[2,1])
    }
    if(nrow(pairwise_matrix) >= 2 && ncol(pairwise_matrix) >= 2 && !is.na(pairwise_matrix[2,2])) {
      comparisons <- c(comparisons, "Imputed (no intermediate) vs Imputed (with intermediate)")
      p_values <- c(p_values, pairwise_matrix[2,2])
    }

    if(length(comparisons) > 0) {
      pairwise_df <- data.frame(
        Comparison = unlist(comparisons),
        P_Value = unlist(p_values),
        Significant = ifelse(unlist(p_values) < 0.05, "Yes", "No")
      )

      kable(pairwise_df, digits = 4, caption = "Pairwise comparisons (Bonferroni adjusted)")
    }
  }
} else {
  cat("Note: ANOVA test could not be performed due to insufficient groups with valid data.\n")
}

# Also do simplified tests for backward compatibility
valid_simple <- changes_df |> filter(!is.na(change_baseline_to_outcome))
if(nrow(valid_simple) > 0 && length(unique(valid_simple$outcome_was_imputed)) == 2) {
  t_test <- t.test(change_baseline_to_outcome ~ outcome_was_imputed, data = valid_simple)

  observed_vals <- valid_simple$change_baseline_to_outcome[!valid_simple$outcome_was_imputed]
  imputed_vals <- valid_simple$change_baseline_to_outcome[valid_simple$outcome_was_imputed]

  if(length(observed_vals) > 0 && length(imputed_vals) > 0) {
    ks_test <- ks.test(imputed_vals, observed_vals)
  } else {
    ks_test <- NULL
  }
} else {
  t_test <- NULL
  ks_test <- NULL
}
```
